---
title: "Hemorrhage Biopsies: Gene module analysis"
output: html_document
author: Naresh Doni Jayavelu
---

&nbsp;

&nbsp;

&nbsp;



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

&nbsp;

&nbsp;

&nbsp;

## Load all necessary libraries

```{r loadLibraries, message=FALSE, warning=FALSE, echo=FALSE}
library(Seurat)
library(patchwork)
# library(monocle3)
# library(scater)
library(stringr)
library(ggpubr)
library(gridExtra)
library(car)
library(lme4)
# library(garnett)
library(ggrepel)
library(ggrastr)
library(dplyr)
library(tidyverse)
library(presto)
library(kableExtra)
library(liana)
library(monocle3)
library(SeuratWrappers)
library(rstatix)
library(clinfun)  # For Jonckheere-Terpstra test
library(openxlsx)


options(stringsAsFactors = FALSE)
set.seed(100)

Csparse_validate = "CsparseMatrix_validate"

```

&nbsp;

&nbsp;

&nbsp;

### Set up the directories
```{r set_up_directories, cache = TRUE}
baseDir <- "/Users/nareshdonijayavelu/Downloads/Projects_Naresh/Hemorrhage_Biopsies"
outDir <- file.path(baseDir, "output")
processed_dataDir <- file.path(baseDir, "processed_data")
raw_dataDir <- file.path(baseDir, "raw_data/GSE266873_RAW")



```

&nbsp;

&nbsp;

&nbsp;

### Read processed seurat object data and DEGs data

```{r Read processed seurat object data}

# read the seurat object
load(file = file.path(processed_dataDir, "seurat_obj_filtered_CCA_integrated_clusters_merged_250221.rda")) # seurat_obj_filtered_CCA_integrated_v3

seurat_obj_filtered_CCA_integrated_v3@meta.data <- seurat_obj_filtered_CCA_integrated_v3@meta.data %>%
  rownames_to_column(var = "cellbc") %>%
  separate(orig.ident, into = c("GEO_sampleid", "sample_number", "sample_number2", "temp", "time"), sep = "_", remove = FALSE) %>%
  mutate(timePoint = case_when(
    time %in% c("2h", "3h") ~ "G1",
    time %in% c("12h") ~ "G2",
    time %in% c("26h", "28h", "32h") ~ "G3",
    TRUE ~ "Other" # Default case for unmatched values
  )) %>%
  column_to_rownames(var = "cellbc")


# read DEGs list
load(file = file.path(processed_dataDir, "DEGs_cluster_timePoints.rda")) # DEGs_cluster_timePoints



```



&nbsp;

&nbsp;

&nbsp;

### Identify gene modules within each cell type

```{r Identify gene modules}

# deg_list = DEGs_cluster_timePoints
# cluster_col = "new_cluster_names"
# resolution = 1e-3

run_gene_modules_per_cluster <- function(seurat_obj, deg_list, cluster_col = "new_cluster_names", resolution = 1e-2) {
 
  # Ensure DEG list has required columns
  if (!all(c("gene_name", "cell_type", "p_val_adj", "avg_log2FC", "pct.1", "pct.2") %in% colnames(deg_list))) {
    stop("DEG list must contain 'gene_name', 'cell_type', 'p_val_adj', 'avg_log2FC', pct.1' and 'pct.2' columns.")
  }
  
  # Initialize an empty list to store results
  all_results <- list()
  
  # Get unique clusters in Seurat object
  unique_clusters <- unique(seurat_obj[[cluster_col]][, 1])
  
  for (cluster in unique_clusters) {
    message(paste("Processing cluster:", cluster))
    
    # Get DEGs for the current cluster
    cluster_degs <- deg_list %>%
      filter(cell_type == cluster, p_val_adj < 0.01, abs(avg_log2FC) > 0.25) %>%
      filter(pct.1 > 0.1 | pct.2 > 0.1) %>%
      distinct(gene_name, .keep_all = TRUE) # Adjust thresholds as needed
    
    # Get gene names and filter only existing genes in Seurat object
    deg_genes <- intersect(cluster_degs$gene_name, rownames(seurat_obj))
    
    # Check if any DEGs are found
    if (length(deg_genes) < 20) {  # Check if enough genes are present
      message(paste("Skipping cluster", cluster, "due to insufficient DEGs (", length(deg_genes), " genes)."))
      next
    }
  
    
    # Subset Seurat object for the cluster
    cluster_cells <- WhichCells(seurat_obj, ident = cluster)
    seurat_subset <- subset(seurat_obj, cells = cluster_cells)
    
    # Convert to Monocle3 CDS
    cds <- as.cell_data_set(seurat_subset)
    
    # Subset CDS to only include user-specified DEGs
    cds <- cds[deg_genes, ]
    
    if (nrow(cds) == 0) {
      message(paste("Skipping cluster", cluster, "as no DEGs are expressed."))
      next
    }
    
    # Preprocess and learn graph structure
    cds <- preprocess_cds(cds, num_dim = 50)
    cds <- reduce_dimension(cds, reduction_method = "UMAP")
    cds <- cluster_cells(cds)
    # cds@clusters$UMAP$partitions <- as.factor(rep(1, length(cds@clusters$UMAP$clusters)))  # Assign a partition
    colData(cds)$assigned_cell_type <- as.character(partitions(cds))
    
    # Find gene modules
    gene_module_tbl <- find_gene_modules(cds, resolution = resolution, random_seed = 100, cores = 4)
    
    # Add cluster information to results
    gene_module_tbl$cluster <- cluster
    
    # Store results
    all_results[[as.character(cluster)]] <- gene_module_tbl
  }
  
  # Combine all results into one dataframe
  final_results <- bind_rows(all_results)
  
  
  return(final_results)
}

seurat_obj <- seurat_obj_filtered_CCA_integrated_v3 %>% subset(new_cluster_names == "Bcells")

# test it on a single cell type
# gene_module_results <- run_gene_modules_per_cluster(seurat_obj, deg_list = DEGs_cluster_timePoints)

gene_module_results <- run_gene_modules_per_cluster(seurat_obj_filtered_CCA_integrated_v3, 
                                                    deg_list = DEGs_cluster_timePoints, 
                                                    cluster_col = "new_cluster_names", 
                                                    resolution = 1e-2)

# save(gene_module_results, file = file.path(processed_dataDir, "gene_module_results.rda"))
load(file = file.path(processed_dataDir, "gene_module_results.rda")) # gene_module_results

gene_module_definitions <- gene_module_results %>%
  mutate(module_id = paste0(cluster, module)) %>%
  dplyr::rename(gene_name = id)

# save to a csv file
write_csv(x = gene_module_definitions, file = file.path(outDir, "Tables/Gene_module_definitions_for_all_cell_types.csv"))


# 
# tt = gene_module_results %>%
#   distinct(cluster, module, .keep_all = TRUE) %>%
#   group_by(cluster, module) %>%
#   summarise(count = n())

```


&nbsp;

&nbsp;

&nbsp;

### Compute Seurat module (from Monocle3) scores
```{r module scores, cache = TRUE}

# format gene modules data to compute Seurat modules scores
gene_modules_data <- gene_module_results %>%
  mutate(module_id = paste0(cluster, module)) %>%
  select(gene_name = id,
         module_id) 
  

# Convert gene_modules_data into a list of gene sets
gene_modules_list <- gene_modules_data %>%
  split(.$module_id) %>%                # Group by module_id
  lapply(function(x) x$gene_name)        # Extract gene names for each module


# Compute module scores
seurat_obj_filtered_CCA_integrated_v4 <- AddModuleScore(
  object = seurat_obj_filtered_CCA_integrated_v3,
  features = gene_modules_list,
  name = "ModuleScore"
)

# change the colnames of default module scores to actual modules
# Get the module IDs
module_ids <- names(gene_modules_list)

# Rename columns in metadata
names(seurat_obj_filtered_CCA_integrated_v4@meta.data)[
  (ncol(seurat_obj_filtered_CCA_integrated_v4@meta.data) - length(module_ids) + 1):ncol(seurat_obj_filtered_CCA_integrated_v4@meta.data)
] <- module_ids

meta_data_with_module_scores <- seurat_obj_filtered_CCA_integrated_v4@meta.data %>%
  mutate(timePoint = case_when(
    time %in% c("2h", "3h") ~ "G1",
    time %in% c("12h") ~ "G2",
    time %in% c("26h", "28h", "32h") ~ "G3",
    TRUE ~ "Other" # Default case for unmatched values
  )) %>%
  mutate(timePoint = factor(timePoint, levels = c("G1", "G2", "G3")))
# save(meta_data_with_module_scores, module_ids,  file = file.path(processed_dataDir, "meta_data_with_monocle3_module_scores.rda"))
load(file = file.path(processed_dataDir, "meta_data_with_monocle3_module_scores.rda"))

# save to a csv file
write_csv(x = meta_data_with_module_scores, file = file.path(outDir, "Tables/Module_scores_with_metadata.csv"))


```


&nbsp;

&nbsp;

&nbsp;

### Compute stats, P-values Seurat module (from monocle3) scores
```{r Compute stats1, cache = TRUE}


data = meta_data_with_module_scores
module_cols = module_columns
group_col = "timePoint"
stratify_col = "new_cluster_names"
test_type = "kruskal"
# group_order = NULL
group_order = c("G1", "G2", "G3")
adjust_within_stratify = TRUE


tt = ggplot(data = data_subset, aes(x = timePoint, y = Astrocytes1)) +
  geom_violin()

# functions to compute P-values by comparing module scores across 3 timepoints
compare_module_scores <- function(data, module_cols, group_col, stratify_col, test_type = "kruskal",
                                  group_order = NULL, adjust_within_stratify = TRUE) {
  # Ensure required columns exist
  if (!all(module_cols %in% colnames(data))) stop("Some module columns are missing in the data frame.")
  if (!(group_col %in% colnames(data))) stop("Grouping column is missing in the data frame.")
  if (!(stratify_col %in% colnames(data))) stop("Stratification column is missing in the data frame.")
  if (!test_type %in% c("kruskal", "wilcox", "t_test", "jonckheere")) stop("Invalid test_type. Choose from 'kruskal', 'wilcox', 't_test', or 'jonckheere'.")

  # Ensure group_order is provided for Jonckheere-Terpstra test
  if (test_type == "jonckheere" & is.null(group_order)) {
    stop("For Jonckheere-Terpstra test, please specify the 'group_order' argument.")
  }

  # Initialize results storage
  all_results <- data.frame()

  # Loop over each stratification group
  for (strat_group in unique(data[[stratify_col]])) {
    data_subset <- data %>% filter(.data[[stratify_col]] == strat_group)
    
    results_list <- list()  # Store results temporarily for this stratification group
    
    # restrict modules to a given cell type
    module_cols_subset <- module_cols[stringr::str_detect(module_cols, as.character(strat_group))]

    # Loop over each module column
    for (mod in module_cols_subset) {
      if (test_type == "kruskal") {
        test_res <- kruskal_test(data_subset, formula(paste(mod, "~", group_col)))
        test_res <- test_res %>% mutate(Group1 = "All", Group2 = "All") %>%
          dplyr::rename(P_Value = p)
      } else if (test_type == "jonckheere") {
        data_subset[[group_col]] <- factor(data_subset[[group_col]], levels = group_order, ordered = TRUE)
        test_result <- jonckheere.test(data_subset[[mod]], data_subset[[group_col]], alternative = "two.sided", nperm = 10000)
        test_res <- tibble(method = "Jonckheere-Terpstra", Group1 = "Ordered", Group2 = "Ordered", P_Value = test_result$p.value)
        # test_res <- tibble(method = "Jonckheere-Terpstra", Stratify_Group = strat_group, Group1 = "Ordered", Group2 = "Ordered", P_Value = test_result$p.value)
      } else {
        test_res <- pairwise_test(data_subset, mod, group_col, test_type)
      }

      test_res <- test_res %>% mutate(Module = mod, Stratify_Group = strat_group)
      results_list[[mod]] <- test_res
    }

    # Combine results for this stratify_group
    strat_results <- bind_rows(results_list)
    all_results <- bind_rows(all_results, strat_results)
  }

  # **Apply p-value adjustment**
  if (adjust_within_stratify) {
    # Adjust p-values **within each stratification group**
    all_results <- all_results %>%
      group_by(Stratify_Group) %>%
      mutate(Adjusted_P_Value = p.adjust(P_Value, method = "BH")) %>%
      ungroup()
  } else {
    # Adjust p-values across **all comparisons together**
    all_results <- all_results %>%
      mutate(Adjusted_P_Value = p.adjust(P_Value, method = "BH"))
  }

  return(all_results)
}

# Function for Wilcoxon and t-test pairwise comparisons
pairwise_test <- function(df, mod, group_col, test_type) {
  if (test_type == "wilcox") {
    res <- df %>% wilcox_test(reformulate(group_col, mod))
  } else if (test_type == "t_test") {
    res <- df %>% t_test(reformulate(group_col, mod))
  }
  res %>%
    select(group1, group2, p) %>%
    dplyr::rename(Group1 = group1, Group2 = group2, P_Value = p)
}




# load metadata with module scores
load(file = file.path(processed_dataDir, "meta_data_with_monocle3_module_scores.rda")) # meta_data_with_module_scores, module_ids


################################################################################
### Compute P-values for Seurat module scores comparing across
### 3 timpoints, G1, G2, G3 ######################
################################################################################

module_columns <- module_ids

kruskal_results <-
  compare_module_scores(
    meta_data_with_module_scores,
    module_cols = module_columns,
    group_col = "timePoint",
    stratify_col = "new_cluster_names",
    test_type = "kruskal",
    group_order = c("G1", "G2", "G3"),
    adjust_within_stratify = TRUE
  )


wilcox_results <-
  compare_module_scores(
    meta_data_with_module_scores,
    module_cols = module_columns,
    group_col = "timePoint",
    stratify_col = "new_cluster_names",
    test_type = "wilcox",
    group_order = c("G1", "G2", "G3"),
    adjust_within_stratify = TRUE
  )

ttest_results <-
  compare_module_scores(
    meta_data_with_module_scores,
    module_cols = module_columns,
    group_col = "timePoint",
    stratify_col = "new_cluster_names",
    test_type = "t_test",
    group_order = c("G1", "G2", "G3"),
    adjust_within_stratify = TRUE
  )

Pvalues_SeuratModuleScores_timePoints <- list(kruskal_results = kruskal_results,
                                              wilcox_results = wilcox_results,
                                              ttest_results = ttest_results)

save(Pvalues_SeuratModuleScores_timePoints, file = file.path(processed_dataDir, "Pvalues_SeuratModuleScores_timePoints.rda"))
# load(file = file.path(processed_dataDir, "Pvalues_SeuratModuleScores_timePoints.rda"))

### save to a xlsx file 
# Create a new workbook
wb <- openxlsx::createWorkbook()

# Add data frames as sheets
addWorksheet(wb, "kruskal_results")
writeData(wb, "kruskal_results", Pvalues_SeuratModuleScores_timePoints$kruskal_results)

addWorksheet(wb, "wilcox_results")
writeData(wb, "wilcox_results", Pvalues_SeuratModuleScores_timePoints$wilcox_results)

addWorksheet(wb, "ttest_results")
writeData(wb, "ttest_results", Pvalues_SeuratModuleScores_timePoints$ttest_results)

# Save the workbook
saveWorkbook(wb, file = file.path(outDir, "Tables/Module_scores_by_timepoints_stats.xlsx"), overwrite = TRUE)



```


&nbsp;

&nbsp;

&nbsp;





